$K$-corrections and filter transformations \\ in the ultraviolet, optical, and near infrared

### 2006-08-03 docs/paper/kcorrect.tex
add
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@562 75dff457-8751-0410-ab1a-cd701654ce6b

\begin{abstract}
Template fits to observed galaxy fluxes allow calculation of
$K$-corrections and conversions among observations of galaxies at
various wavelengths. We present a method for creating model-based
template sets given a set of heterogeneous photometric and
spectroscopic galaxy data. Our technique, non-negative matrix
factorization, is akin to principle component analysis (PCA), except
that it is constrained to produce nonnegative templates, it can use a
basis set of models (rather than the delta function basis of PCA), and
it naturally handles uncertainties, missing data, and heterogeneous
data (including broad-band fluxes at various redshifts).  The
particular implementation we present here is suitable for ultraviolet,
optical, and near-infrared observations in the redshift range $0 < z <
1.5$. Since we base our templates on stellar population synthesis
models, the results are intepretable in terms of approximate stellar
masses and star-formation histories. We present templates fit with
this method to data from GALEX, Sloan Digital Sky Survey spectroscopy
and photometry, the Two-Micron All Sky Survey, the Deep Extragalactic
Evolutionary Probe and the Great Observatories Origins Deep Survey.
In addition, we present software for using such data to estimate
$K$-corrections and stellar masses.
\end{abstract}

@@

We note that some of the algorithmic techniques used here may have
applications in other areas of astrophysics. First, nonnegative matrix
factorization (NMF; \citealt{lee00a}), and the extensions to it we
describe here, is a particularly useful variant on principal component
analysis. Second, the nonnegative least squares algorithm of
\citet{sha02a} has the virtue of being extremely simple to implement.
These methods may find other applications in image analysis, spectroscopic
analysis, and model fitting in astrophysics.

We also release the templates in electronic form as well as an
implementation of the methods used to fit the templates to data. This
software {\tt kcorrect v4\_1} is distributed on the World Wide
Web{\footnote {\tt http://cosmo.nyu.edu/blanton/kcorrect/}}. It
consists of a core C library which performs most of the complex and
computationally intensive tasks, plus an IDL library that provides a
high-level interface.  This software is an update of two major earlier
releases ({\tt v1\_16}; \citealt{blanton03b}; {\tt v3\_2}). The
improvement over the previous version is twofold. First, we have
improved the templates such that they successfully fit galaxies in the
restframe UV, as observed by GALEX, DEEP2, and GOODS. Second, because
the templates are now completely model-based, the fits have a physical
interpretation in terms of a star-formation history. The IDL library
also has a number of new useful functions.

@@

From the spectroscopic observations we know that galaxy spectra reside
in a low-dimensional subspace. Principal Component Analysis (PCA) of
galaxy spectra, introduced by \citet{connolly95b} and applied many
times since then, demonstrates that most of the variance in the
distribution of galaxies in spectral space can be explained using a
few templates. This means that, in the linear space of all possible
spectra, galaxies exist in only a small subspace. Therefore, even with
very heterogeneous data, we should be able to determine the properties
of this subspace.

Here we present an approach to combining heterogeneous data in order
to determine the properties of the subspace of galaxy spectra. Rather
than taking the model-free approach used by PCA, we here restrict the
space of possible spectra to those predicted from the high resolution
stellar population synthesis model of \citet{bruzual03a} and the
nebular emission line models of \citet{kewley01a}.  This approach both
constrains the problem appropriately and yields a natural theoretical
interpretation of the results in terms of star-formation histories.

@@

We begin with the simplest case, which is fitting a {\it single}
template to just the SDSS photometric data of the LRG sample
(\citealt{eisenstein01a}). For many of these galaxies, which extend to
$r<19.5$ and are intrinsically red, the $u$-band flux is extremely
poorly measured, so we ignore the $u$-band for all LRGs and only fit
to $g$, $r$, $i$ and $z$ fluxes using the SDSS ``model'' fluxes as
described in \cite{stoughton01a} and elsewhere.
In this case, we fit for the $a_{i0}$ and $b_{0j}$ coefficients 
in Equation \ref{chi2}: we want just a single template, expressed 
by the $b_{0j}$ coefficients, and a single scaling of its 
flux for each galaxy, expressed by $a_{i0}$.

@@

In addition to $K$-corrections, these template fits also provide
physical interpretations of the galaxies, since they correspond to
actual stellar population synthesis models. Indeed, the code outputs
three parameters relevant to the star-formation history: the current
mass of stars in the galaxy; the stellar mass-to-light ratio of the
galaxy; and the fraction of the total amount of star-formation that
has occurred recently. We do not believe that fitting just to
broad-band colors can ever give extremely accurate values for these
quantities. In this section we demonstrate consistency between our
results and those of other groups for the same galaxies.

@@

Finally, we can also calculate measures of the recent star-formation
rate from these fits. Since our fits are to broad-band photometry, and
not based on emission line measurements, we cannot expect to have
detailed information about the very recent star-formation
rate. However, we can try to measure what fraction of the total
star-formation has occurred in, say, the past 1 Gyr:
\begin{equation}
b_{G} = \frac{\int_0^{1\mathrm{~Gyr}} dt\, \mathrm{SFR}(t)}
{\int_0^{a_{\mathrm{Univ}}} dt\, \mathrm{SFR}(t)}
\end{equation}
where the definition here is meant to be similar to the birthrate
parameter $b$ in \citet{kennicutt94a} and references therein.  Note
that here $b_G$ is defined in terms of the total integrated
star-formation history (unlike the stellar masses referred to above)
as is standard for the $b$ parameter. Figure \ref{umr_bg} shows the
relationship between the rest-frame $u-r$ color for SDSS galaxies and
$b_G$. For galaxies with $u-r < 2.5$ (redder than which galaxies are
mostly reddened, edge-on disks) these two parameters follow a simple
relationship:
\begin{equation}
\log_{10} b_g = -0.55 (u-r). 
\end{equation}
Note, however, the large scatter (about 0.3 dex at 1$\sigma$).  The
red dots show the results of using the formula of \citet{hopkins03a}
to infer the current star-formation rate from the $u$-band,
integrating over 1 Gyr, scaling by the integrated stellar mass from
the {\tt kcorrect} determination, and then multiplying by four (which
is the minimum appropriate correction for dust extinction in $u$
according to \citealt{hopkins03a}). Our results roughly scale as those
of \citet{hopkins03a} do.

@@

\subsection{Standard NMF}
The standard non-negative matrix factorization (NMF) problem, as
originally posed by \cite{lee00a}, is to approximate a data
matrix $\XX$ (of size $N_k\times N_n$) as the outer product of two
rank-$N_i$ non-negative matrices $\WW$ and $\HH$. For example, in this
paper, $\XX$ represents the fluxes at $N_n$ different wavelengths for
$N_k$ different galaxies, $\HH$ represents a set of $N_i$ templates to build
each galaxy from, and $\WW$ represents, for each galaxy, the weights
to give each template. This factorization is similar in spirit to
singular value decomposition (SVD) and principal components analysis
(PCA) but crucially, in NMF, $\XX$, $\WW$ and $\HH$ are all
constrained to have non-negative entries. This changes the problem
fundamentally, since no ``cancellation'' of positive and negative
basis functions or coefficients is possible -- all approximation
interactions are strictly additive. The goal, as with SVD or PCA, is
to minmize the squared approximation error (Frobenius norm):
\beqa \label{eq:nmfcost}
\chi^2 &= \|\XX-\WW\HH\|^2\\
%&= \|\XX-\XXh\|^2\\
&= \sum_{kn} \left(\Xkn - \sum_i \Wki\Hin \right)^2
\eeqa
In their seminal papers, \cite{lee00a} show that this approximation
error in non-increasing under the following very simple multiplicative
update rules: 
\beqa \label{eq:wupdate}
\Wki &\leftarrow \Wki \frac{[\XX\HH\T]_{ki}}{[\WW\HH\HH\T]_{ki}}\\
\label{eq:hupdate}
\Hin &\leftarrow \Hin \frac{[\WW\T\XX]_{in}}{[\WW\T\WW\HH]_{in}}
\eeqa
$[M]_{ab}$ denotes the $(ab)$ element of the matrix valued expression
$M$. $\HH\T$ is the transpose of matrix $\HH$.

>>>>

\begin{abstract}
Template fits to observed galaxy fluxes allow calculation of
$K$-corrections and conversions among observations of galaxies at
various wavelengths. We present a method for creating model-based
template sets given a set of heterogeneous photometric and
spectroscopic galaxy data. Our technique, non-negative matrix
factorization, is akin to principle component analysis (PCA), except
that it is constrained to produce nonnegative templates, it can use a
basis set of models (rather than the delta function basis of PCA), and
it naturally handles uncertainties, missing data, and heterogeneous
data (including broad-band fluxes at various redshifts).  The
particular implementation we present here is suitable for ultraviolet,
optical, and near-infrared observations in the redshift range $0 < z <
1.5$. Since we base our templates on stellar population synthesis
models, the results are intepretable in terms of approximate stellar
masses and star-formation histories. We present templates fit with
this method to data from GALEX, Sloan Digital Sky Survey spectroscopy
and photometry, the Two-Micron All Sky Survey, the Deep Extragalactic
Evolutionary Probe and the Great Observatories Origins Deep Survey.
In addition, we present software for using such data to estimate
$K$-corrections.
\end{abstract}

@@

We note that some of the algorithmic techniques used here may have
applications in other areas of astrophysics. First, nonnegative matrix
factorization (NMF; \citealt{lee00a}), and the extensions to it we
describe here, is a particularly useful variant on principal
components
analysis (PCA). Second, the nonnegative least squares algorithm of
\citet{sha02a} has the virtue of being extremely simple to implement.
These methods may find other applications in image analysis, spectroscopic
analysis, and model fitting in astrophysics.

We also release the templates in electronic form and an implementation
of the methods used to fit the templates to data. This software {\tt
kcorrect v4\_1} is distributed on the World Wide Web{\footnote {\tt
http://cosmo.nyu.edu/blanton/kcorrect/}}. It consists of a core C
library which performs most of the complex and computationally
intensive tasks, plus an IDL library that provides a high-level
interface.  This software is an update of two major earlier releases
({\tt v1\_16}; \citealt{blanton03b}; {\tt v3\_2}). The improvement
over the previous version is twofold. First, we have improved the
templates such that they successfully fit galaxies in the restframe
UV, as observed by GALEX, DEEP2, and GOODS. Second, because the
templates are now completely model-based, the fits have a physical
interpretation in terms of a star-formation history. The IDL library
also has a number of new useful functions.

@@

From the spectroscopic observations we know that galaxy spectra reside
in a low-dimensional subspace of all possible spectra. Principal
Component Analysis (PCA) of galaxy spectra, introduced by
\citet{connolly95b} and applied many times since then, demonstrates
that most of the variance in the distribution of galaxies in spectral
space can be explained using a few templates. This means that, in the
linear space of all possible spectra, galaxies exist in only a small
subspace. Therefore, even with very heterogeneous data, we should be
able to determine the properties of this subspace.

Here we present an approach to combining heterogeneous data in order
to determine the properties of the subspace of galaxy spectra. Rather
than taking the model-free approach used by PCA, we here restrict the
space of possible spectra to those predicted from the high resolution
stellar population synthesis model of \citet{bruzual03a} and the
nebular emission line models of \citet{kewley01a}.  This approach both
constrains the problem appropriately and yields a natural theoretical
interpretation of the results in terms of star-formation
histories. Generalizing this approach to use other models of emission
from stars and gas is straightforward.

@@

We begin with the simplest case, which is fitting a {\it single}
template to just the SDSS photometric data of the LRG sample
(\citealt{eisenstein01a}). For many of these galaxies, which extend to
$r<19.5$ and are intrinsically red, the $u$-band flux is extremely
poorly measured, so we ignore the $u$-band for all LRGs and only fit
to $g$, $r$, $i$ and $z$ fluxes using the SDSS ``model'' fluxes as
described in \cite{stoughton02a} and elsewhere.
In this case, we fit for the $a_{i0}$ and $b_{0j}$ coefficients 
in Equation \ref{chi2}: we want just a single template, expressed 
by the $b_{0j}$ coefficients, and a single scaling of its 
flux for each galaxy, expressed by $a_{i0}$.

@@

In addition to $K$-corrections, these template fits also provide
physical interpretations of the galaxies, since they correspond to
actual stellar population synthesis models. Indeed, the code outputs
three parameters relevant to the star-formation history: the current
mass of stars in the galaxy; the stellar mass-to-light ratio of the
galaxy; and the fraction of the total amount of star-formation that
has occurred recently. We do not believe that fitting just to
broad-band colors can ever give extremely accurate values for these
quantities, but to evaluate what {\tt kcorrect} does in this regard we
here examine the consistency between our results and those of other
groups for the same galaxies.

@@

Finally, we can also calculate measures of the recent star-formation
rate from these fits. Since our fits are to broad-band photometry, and
not based on emission line measurements, we cannot expect to have
detailed information about the very recent star-formation
rate. However, we can try to measure what fraction of the total
star-formation has occurred in, say, the past 300 Myr:
\begin{equation}
b_{300} = \frac{\int_0^{300\mathrm{~Myr}} dt\, \mathrm{SFR}(t)}
{\int_0^{a_{\mathrm{Univ}}} dt\, \mathrm{SFR}(t)}
\end{equation}
where the definition here is meant to be similar to the birthrate
parameter $b$ in \citet{kennicutt94a} and references therein.  Note
that here $b_{300}$ is defined in terms of the total integrated
star-formation history (unlike the stellar masses referred to above)
as is standard for the $b$ parameter. Figure \ref{umr_bg} shows the
relationship between the rest-frame $u-r$ color for SDSS galaxies and
$b_{300}$. For our results, the median relationship is close to:
\begin{equation}
\log_{10} b_{300} = 0.2 - 1.3 (u-r). 
\end{equation}
Note, however, the large scatter (about 0.3 dex at 1$\sigma$).  The
overlaid points show the results of using the formula of \citet{hopkins03a}
to infer the current star-formation rate from the $u$-band,
integrating over 300 Myr, scaling by the integrated stellar mass from
the {\tt kcorrect} determination, and then correcting for the internal
$u$-band dust extinction inferred by the {\tt kcorrect} fits (which
turns out to be factors ranging from 3--10, similar to the factors
used by \citealt{hopkins03a}).  Our results roughly scale as those of
\citet{hopkins03a} do, but with lower star-formation rates by a factor
of five or so. Since \citet{hopkins03a} calibrate their star-formation
rates to a number of other indicators, we conclude that either {\tt
kcorrect} probably underestimates the recent star-formation rate.

@@

\subsection{Standard NMF}
The standard non-negative matrix factorization (NMF) problem, as
originally posed by \cite{lee00a}, is to approximate a data matrix
$\XX$ (of size $N_k\times N_n$) as the outer product of two rank-$N_i$
non-negative matrices $\WW$ and $\HH$. For example, in this paper,
$\XX$ represents the fluxes at $N_n$ different wavelengths for $N_k$
different galaxies, $\HH$ represents a set of $N_i$ templates to build
each galaxy from, and $\WW$ represents, for each galaxy, the weights
to give each template. This factorization is similar in spirit to
singular value decomposition (SVD) and PCA, but crucially, in NMF,
$\XX$, $\WW$ and $\HH$ are all constrained to have non-negative
entries. This changes the problem fundamentally, since no
``cancellation'' of positive and negative basis functions or
coefficients is possible -- all approximation interactions are
strictly additive. The goal, as with SVD or PCA, is to minmize the
squared approximation error (Frobenius norm): \beqa \label{eq:nmfcost}
\chi^2 &= \|\XX-\WW\HH\|^2\\
%&= \|\XX-\XXh\|^2\\
&= \sum_{kn} \left(\Xkn - \sum_i \Wki\Hin \right)^2
\eeqa
In their seminal papers, \cite{lee00a} show that this approximation
error in non-increasing under the following very simple multiplicative
update rules: 
\beqa \label{eq:wupdate}
\Wki &\leftarrow \Wki \frac{[\XX\HH\T]_{ki}}{[\WW\HH\HH\T]_{ki}}\\
\label{eq:hupdate}
\Hin &\leftarrow \Hin \frac{[\WW\T\XX]_{in}}{[\WW\T\WW\HH]_{in}}
\eeqa
$[M]_{ab}$ denotes the $(ab)$ element of the matrix valued expression
$M$. $\HH\T$ is the transpose of matrix $\HH$.

### 2006-08-03 docs/paper/kcorrect_figures.tex
paper changes
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@560 75dff457-8751-0410-ab1a-cd701654ce6b
add
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@562 75dff457-8751-0410-ab1a-cd701654ce6b

\clearpage
\stepcounter{thefigs}
\begin{figure}
\figurenum{\fignum}
\plotone{umr_bg.ps}
\caption{\label{umr_bg} Fraction of the total star-formation that has
	occurred in the previous billion years $b_G$, as a function of restframe
	$u-r$ color, for SDSS galaxies. The greyscale is the conditional
	distribution of $b_G$ on $u-r$. The lines are the 10\%, 25\%, 50\%,
	75\% and 90\% quantiles. For $u-r < 2.5$, the median relationship
	follows the simple form listed in the figure. Galaxies with $u-r >
	2.5$ are often highly reddened star-forming galaxies.}
\end{figure}

>>>>

\clearpage
\stepcounter{thefigs}
\begin{figure}
\figurenum{\fignum}
\plotone{umr_bg.ps}
\caption{\label{umr_bg} Fraction of the total star-formation that has
	occurred in the previous 300 Myr $b_{300}$, as a function of
	restframe $u-r$ color, for SDSS galaxies. The greyscale is the
	conditional distribution of $b_{300}$ on $u-r$. The lines are the 10\%,
	25\%, 50\%, 75\% and 90\% quantiles.  For comparison we show the
	results of using the scaling of \citet{hopkins03a}.}
\end{figure}
